Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-09 latex/main.tex
Updates from Overleaf

To evaluate the concordance of the methods in
identifying the Pango lineages of recombinant parents and locating recombination breakpoints,
we focus on the Pango Xs associated with the Type I recombination events (Table~\ref{tab:pango_x_lineages}),
using the data from the Pango designation community (https://github.com/cov-lineages/pango-designation/) as a common comparison.
For RecombinHunt and CovRecomb, we summarized detection results for these Pango Xs, if available,
reported in the original studies\cite{Alfonsi2024,Li2024CovRecomb} (STAR Methods).
For each method, we determined the Pango Xs for which the inferred parental lineages and breakpoint intervals
are concordant with those identified by the community (STAR Methods).
We found that the methods performed well overall,
with comparable concordance rates (Table~\ref{tab:method_concordance}),
indicating that sc2ts can characterize recombinants as well as the state-of-the-art methods.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=1\linewidth]{recombinant_evidence.pdf}
    \caption{Evidence for recombination in the ARG. All plots show ``robust'' sc2ts recombination nodes, colored according
to their classification by rebar (recombinant: purple, non-recombinant: gray).
(A) Upper: jittered scatterplot comparing number of mutations when recombination allowed ($y$: right axis)
versus an alternative copying path disallowing recombination ($x$: top axis);
recombination nodes in Table \ref{tab:pango_x_lineages} are labelled starting with `X' and
those identified by Jackson et al.\cite{Jackson2021} are prefixed with a `J', e.g.
Jackson Group B (JGB) and Jackson Singleton recombinant 1 (JS1). 
Point sizes reflect number of descendants.
Lower: recombination nodes classified by number of mutations averted by allowing recombination (i.e. $x$-$y$).
(B) Distribution of time to the MRCA of the parents of each robust recombination node.
(C) Distribution of the Pango lineage distance between parents.
(D) Copying patterns leading to XA, XBB, XBB.1, and the recombinant origin of XZ+XAC+XAD+XAE+XAP;
each recombination node is a mix of variants matching the left parent (green), the right parent (blue), and de-novo mutations (gold).}
    \label{fig:averted_mutations}
\end{figure}

@@

\subsection{Analysis of deletions}\label{sec:supplement_dels}
% Referenced by the main Results.
Li et al.\cite{LiStephens2003} performed a large-scale analysis of indels 
using a global GISAID dataset (sampled up to August, 2022). 
We examined the major indels reported in Table S1 of that study 
which have a minimum frequency of 1\% 
(sequence count divided by 9,149,680 filtered sequences).
Most of these indels (all deletions) are encoded in the ARG (Table~\ref{tab:indels}) 
and have high frequencies in the ARG
(sequence count divided by the total number of samples), 
except a 9-base deletion at 11283, a 9-based deletion at 21987, and a 9-base deletion at 686.
During site remapping, the 9-base deletion at 11283 was split up into 
a 5-base deletion at 11283 (count of 339867) and 
a 4-base deletion at 11288 (count of 871782).
The 686 deletion occurs below our threshold for inclusion into the ARG.

% See https://github.com/jeromekelleher/sc2ts-paper/issues/472
\begin{table}
\caption{Major indels.}
\label{tab:indels}
\begin{center}
\begin{tabular}{rrcrrrr}
\toprule 
& \multicolumn{4}{c}{Li et al. (2023)} & \multicolumn{2}{c}{Sc2ts ARG} \\
\cmidrule(lr){1-5} \cmidrule(lr){6-7}
Start & Length & Region & Count & Frequency (\%) & Count & Frequency (\%)\\
\midrule
28362 & 9 & N & 3939164 & 43.0525 & 871774 & 35.1216 \\
28271 & 1 & Intergenic & 3853692 & 42.1183 & 1118363 & 45.0561 \\
21765 & 6 & S & 3179887 & 34.7541 & 339681 & 13.6849 \\
11288 & 9 & ORF1ab-nsp6 & 3040576 & 33.2315 & 290408 & 11.6998 \\
28248 & 6 & ORF8 & 3026186 & 33.0742 & 1114669 & 44.9073 \\
22029 & 6 & S & 3010665 & 32.9046 & 1112410 & 44.8163 \\
11283 & 9 & ORF1ab-nsp6 & 2011934 & 21.9891 & 40 & 0.0016 \\
6513 & 3 & ORF1ab-nsp3 & 1984683 & 21.6913 & 340765 & 13.7286 \\
21633 & 9 & S & 1941962 & 21.2244 & 532072 & 21.4359 \\
21987 & 9 & S & 1918094 & 20.9635 & N/A & N/A \\
22194 & 3 & S & 1833050 & 20.0340 & 312706 & 12.5982 \\
21991 & 3 & S & 912944 & 9.9779 & 288831 & 11.6363 \\
686 & 9 & ORF1ab-nsp1 & 115450 & 1.2618 & N/A & N/A \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

>>>>

To evaluate the concordance of the methods in
identifying the Pango lineages of recombinant parents and locating recombination breakpoints,
we focus on the Pango Xs associated with the Type I recombination events (Table~\ref{tab:pango_x_lineages}),
using the data from the Pango designation community (https://github.com/cov-lineages/pango-designation/) as a common comparison.
For RecombinHunt and CovRecomb, we summarized the detection results for these Pango X lineages (if available)
reported in the original studies\cite{Alfonsi2024,Li2024CovRecomb} (STAR Methods).
For each method, we determined the Pango X lineages
for which the inferred parental lineages and breakpoint intervals
are concordant with those identified by the community (STAR Methods).
We found that the methods performed well overall,
with comparable concordance rates (Table~\ref{tab:method_concordance}),
indicating that sc2ts can characterize recombinants as well as the state-of-the-art methods.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=1\linewidth]{recombinant_evidence.pdf}
    \caption{Evidence for recombination in the ARG. All plots show ``robust'' sc2ts recombination nodes, colored according
to their classification by rebar (recombinant: purple, non-recombinant: gray).
(A) Upper: jittered scatterplot comparing number of mutations when recombination allowed ($y$: right axis)
versus an alternative copying path disallowing recombination ($x$: top axis);
recombination nodes in Table \ref{tab:pango_x_lineages} are labeled starting with `X' and
those identified by Jackson et al.\cite{Jackson2021} are prefixed with a `J', e.g.
Jackson Group B (JGB) and Jackson Singleton recombinant 1 (JS1). 
Point sizes reflect number of descendants.
Lower: recombination nodes classified by number of mutations averted by allowing recombination (i.e. $x$-$y$).
(B) Distribution of time to the MRCA of the parents of each robust recombination node.
(C) Distribution of the Pango lineage distance between parents.
(D) Copying patterns leading to XA, XBB, XBB.1, and the recombinant origin of XZ+XAC+XAD+XAE+XAP;
each recombination node is a mix of variants matching the left parent (green), the right parent (blue), and de-novo mutations (gold).}
    \label{fig:averted_mutations}
\end{figure}

@@

\subsection{Analysis of deletions}\label{sec:supplement_dels}

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
% Referenced by the main Results.
We found that the ARG captures highly prevalent deletions in SARS-CoV-2 (Table~\ref{tab:major_dels}).
Li et al.\cite{Li2023indels} performed a large-scale analysis of indels
using a global GISAID dataset (sampled up to August, 2022).
We examined the major indels reported in Table S1 of that study
which have a minimum frequency of 1\% (sequence count divided by 9,149,680 filtered sequences).
Most of these indels (all deletions) are encoded in the ARG (Table~\ref{tab:indels})
and have high frequencies in the ARG (sequence count divided by the total number of samples),
except a 9-base deletion at 11283, a 9-based deletion at 21987, and a 9-base deletion at 686.
During site remapping, the 9-base deletion at 11283 was split up into
a 5-base deletion at 11283 (count of 339867) and a 4-base deletion at 11288 (count of 871782).

\begin{table}
\caption{
Major deletion events in the ARG.
These events are observed in at least 10,000 samples encoded in the ARG.
The counts and frequencies of these events from Li et al.\cite{Li2023indels} are also shown.
}
\label{tab:major_dels}
\begin{tabular}{rrcrcrrrr}
\centering
\toprule & \multicolumn{6}{c}{Sc2ts ARG} & \multicolumn{2}{c}{Li et al. (2023)} \\
\cmidrule(lr){1-7} \cmidrule(lr){8-9}
Start & Length & Region & Node & Pango & Count & Frequency (\%) & Count & Frequency (\%)\\
\midrule
28271 & 1 & non-coding & 1436808 & B.1 & 1118363 & 45.0561 & 3853692 & 42.1183 \\
28248 & 6 & ORF8 & 220186 & B.1.617.2 & 1114669 & 44.9073 & 3026186 & 33.0742 \\
22029 & 6 & S / S1 - NTD & 200039 & B.1.617.2 & 1112410 & 44.8163 & 3010665 & 32.9046 \\
11288 & 4 & ORF1AB / NSP6 & 1436802 & B.1.1.529 & 871782 & 35.1220 & N/A & N/A \\
28362 & 9 & N & 1436802 & B.1.1.529 & 871774 & 35.1216 & 3939164 & 43.0525 \\
21633 & 9 & S / S1 - NTD & 822854 & BA.2 & 532072 & 21.4359 & 1941962 & 21.2244 \\
11292 & 5 & ORF1AB / NSP6 & 822854 & BA.2 & 531894 & 21.4287 & N/A & N/A \\
6513 & 3 & ORF1AB / NSP3 & 851246 & BA.1 & 340765 & 13.7286 & 1984683 & 21.6913 \\
11283 & 5 & ORF1AB / NSP6 & 851246 & BA.1 & 339867 & 13.6924 & 8360 & 0.0914 \\
21765 & 6 & S / S1 - NTD & 851246 & BA.1 & 339681 & 13.6849 & 3179887 & 34.7541 \\
21988 & 8 & S / S1 - NTD & 851246 & BA.1 & 339651 & 13.6837 & 590 & 0.0064 \\
22194 & 3 & S / S1 - NTD & 851246 & BA.1 & 312706 & 12.5982 & 1833050 & 20.0340 \\
28271 & 1 & non-coding & 86456 & B.1.1.7 & 290446 & 11.7014 & 3853692 & 42.1183 \\
11288 & 9 & ORF1AB / NSP6 & 86456 & B.1.1.7 & 290408 & 11.6998 & 3040576 & 33.2315 \\
21991 & 3 & S / S1 - NTD & 86456 & B.1.1.7 & 288831 & 11.6363 & 912944 & 9.9779 \\
21765 & 6 & S / S1 - NTD & 86456 & B.1.1.7 & 290438 & 11.7010 & 3179887 & 34.7541 \\
21765 & 6 & S / S1 - NTD & 1265302 & BA.4 & 190425 & 7.6718 & 3179887 & 34.7541 \\
28361 & 1 & N & 1246012 & BA.5.2.1 & 20822 & 0.8389 & 287 & 0.0031 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\caption{
Deletion events associated with the Alpha variant in the ARG.
These events occur above node 86456, which represents the origin of Alpha.
}
\label{tab:alpha_dels}
\centering
\begin{tabular}{rrrrr}
\toprule
Start & Length & Region & Count & Frequency (\%)\\
\midrule
28271 & 1 & non-coding & 290446 & 11.7014 \\
11288 & 9 & ORF1AB / NSP6 & 290408 & 11.6998 \\
21991 & 3 & S / S1 - NTD & 288831 & 11.6363 \\
21765 & 6 & S / S1 - NTD & 290438 & 11.7010 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\caption{
Highly recurrent deletions.
}
\label{tab:rec_dels}
\centering
\begin{tabular}{rrcrr}
\toprule
Start & Length & Region & Occurrences \\
\midrule
29781 & 1 & non-coding & 12713 \\
29762 & 1 & non-coding & 11356 \\
29779 & 1 & non-coding & 10925 \\
29769 & 1 & non-coding & 10783 \\
29774 & 1 & non-coding & 10707 \\
29700 & 1 & non-coding & 8543 \\
29614 & 1 & ORF10 & 5041 \\
29555 & 1 & non-coding & 4816 \\
29543 & 1 & non-coding & 4806 \\
29541 & 1 & non-coding & 4804 \\
203 & 1 & non-coding & 3602 \\
222 & 1 & non-coding & 3441 \\
22013 & 6 & S / S1 - NTD & 2111 \\
21991 & 3 & S / S1 - NTD & 2087 \\
28254 & 1 & ORF8 & 1967 \\
27556 & 64 & ORF7a & 1281 \\
27579 & 3 & ORF7a & 1014 \\
28093 & 1 & ORF8 & 898 \\
21765 & 6 & S / S1 - NTD & 478 \\
27532 & 1 & ORF7a & 403 \\
\bottomrule
\end{tabular}
\end{table}
